Efficient Verified Implementation of Introsort and Pdqsort

### 2020-02-13 papers/sorting/main.tex
pdqsort paper

\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{balance}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

@@

@@

@@

\theoremstyle{definition}
\newtheorem{example}{Example}[section]

@@

\title{Efficient Verified Implementation of Introsort}

\author{Peter Lammich}
\email{peter.lammich@machester.ac.uk}
\affiliation{%
  \institution{The University of Manchester, UK}
}
%
% \authorinfo{Peter Lammich}
%            {The University of Manchester, UK}
%            {peter.lammich@machester.ac.uk}

\begin{abstract}
  Sorting algorithm are an important part of most standard libraries, and
  both, their correctness and efficiency is crucial for many applications.
%   There has been extensive research on improving sorting algorithms for
%   both, the general case of sorting arbitrary objects wrt.\ a weak ordering,
%   or special cases like sorting integers or strings.

  As generic sorting algorithm, the GNU C++ Standard Library implements the Introsort algorithm,
  a combination of quicksort, heapsort, and insertion sort.

  We verify this algorithm in the Isabelle LLVM verification framework,
  including most of the optimizations from the GNU implementation.
  On an extensive benchmark set, our verified algorithm performs on par with the original.
\end{abstract}

@@

@@

\section{Introduction}
  The goal of this paper is to provide a case study to show that the verification of
  state-of-the art implementations of medium-complex algorithms is feasible.
  We chose Musser's introspective sorting algorithm (Introsort)~\cite{Muss97}, and its
  implementation in the GNU C++ Library (libstdc++)~\cite{libstdc++}.
  Using the Isabelle-LLVM verification framework~\cite{La19-llvm}, we verify the same algorithm,
  including most of the optimizations from libstdc++. The result is a verified sorting algorithm, which performs
  on-par with the original implementation on an extensive set of benchmarks.

  Sorting algorithms are an important part of any standard library. The Introsort algorithm,
  being a combination of quicksort, heapsort, and insertion sort, is far from being trivial, but still
  has a manageable complexity. Finally, libstdc++ provides a state-of-the art implementation with several
  non-trivial but crucial optimizations. This makes Introsort a good candidate for our case study.

  Moreover, sorting algorithms in standard libraries have not always been correct.
  The Timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
  Also, many real-world mergesort implementations suffered from an overflow bug~\cite{bs_flaw_blogpost}.
  Finally, LLVM's libc++~\cite{libc++} implements a different quicksort based sorting algorithm. While it may be functionally correct,
  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by October 2019.}.

  To the best of our knowledge, this paper provides the first verification of the Introsort algorithm.
  This algorithm and two Java sorting algorithms~\cite{GRBBH15,BSSU17} are the only verified "real-world" implementations
  of sorting algorithms we are aware of. On the technical side, this paper provides a case study of using stepwise refinement
  for the design of verified algorithms.
  Moreover, we report on two extensions to the Isabelle-LLVM framework, to handle nested container data structures and to automatically
  generate C-header files to interface the generated code. These were required for our benchmark programs, which sort arrays of arrays of characters,
  and interface the generated LLVM code from C++.

  The complete Isabelle/HOL formalization and the benchmarks are submitted as artefact accompanying this paper.

@@

\section{The Introsort Algorithm}
  The Introsort algorithm by Musser~\cite{Muss97} is a sorting algorithm
  that combines the good average-case runtime of quicksort~\cite{Hoare61} with the worst-case complexity of heapsort~\cite{Will64}.
  The basic idea is to use quicksort as main sorting algorithm, but switch to heapsort when the recursion
  depth exceeds a given limit, usually $2\lfloor\log_2 n\rfloor$ for $n$ elements.

  Introsort is the default generic sorting algorithm in the GNU C++ Library~\cite{libstdc++}. Alg.~\ref{alg:introsort} displays its pseudo-code.
  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{partial\_introsort}{$a,l,h,d$}
      \If{$h-l>\textrm{threshold}$}
        \If{$d=0$}
          \State{\Call{heapsort}{$a,l,h$}}
        \Else
          \State{$m\gets\Call{partition\_pivot}{a,l,h}$}
          \State{\Call{partial\_introsort}{$a,l,m,d-1$}}
          \State{\Call{partial\_introsort}{$a,m,h,d-1$}}
        \EndIf
      \EndIf
    \EndProcedure

    \Procedure{introsort}{$a,l,h$}
      \If{$h-l>1$}
        \State{\Call{partial\_introsort}{$a,l,h,2\lfloor\log_2(h-l)\rfloor$}}
        \State{\Call{final\_insort}{$a,l,h$}}
      \EndIf
    \EndProcedure
  \end{algorithmic}
  \caption{Introsort}\label{alg:introsort}
  \end{algorithm}
  The \textsc{introsort} function sorts the slice $[l{..}{<}h]$ of an array $a$. For this, it first
  calls \textsc{partial\_introsort} to partially sort the array, i.e., all elements are no more than $\textrm{threshold}$ from their final position,
  and then finishes the sorting by invoking insertion sort. The \textsc{partial\_introsort} procedure implements a standard quicksort, that switches
  to heapsort when the recursion depth gets greater than $d$, and stops when the partition size becomes less than or equal to $\textrm{threshold}$.
  Note that the actual implementation in the GNU C++ Library contains a manual tail-call optimization,
  replacing the second recursive call to \textsc{partial\_introsort} by looping. For our formalization, we chose to not implement this
  optimization, in particular due to the fact that the LLVM optimizer recognizes and eliminates this tail call automatically.

  \subsection{Partitioning}
  The \textsc{partition\_pivot($a,l,h$)} algorithm must re-arrange the elements in $a$ and return a \emph{pivot} index $m\in[l{<}{..}{<}h]$, such
  that all elements $a[l{..}{<}m]$ are less than or equal to $a[m]$, and all elements $a[m{..}{<}h]$ are greater than or equal to $a[m]$.
  Note that the performance of quicksort crucially depends on pivot selection: Ideally, the pivot index should be close to $(h-l)/2$,
  i.e., the pivot element should be close to the median. The algorithm implemented by the GNU C++ Library is sketched in Alg.~\ref{alg:partition}.
  It selects the median of the three elements $a[l+1]$, $a[\lfloor (h-l)/2 \rfloor]$, and $a[h-1]$ as pivot (l.~\ref{line:partition:pivot_selection}).
  It then swaps the selected median element to the front $a[l]$, and invokes the standard Hoare partitioning algorithm,
  which searches for pairs of mismatched elements simultaneously from the start and end of the array, and swaps them.
  The chosen pivot element remains at position $a[l]$.
  \begin{algorithm}
  \begin{algorithmic}[1]
    \Function{partition}{$a,l,h,p$}
      \While{True}
        \While{$a[l]<a[p]$} ${+}{+}l$ \EndWhile \label{partition:nextl_loop}
        \State{${-}{-}h$}
        \While{$a[p]<a[h]$} ${-}{-}h$ \EndWhile \label{partition:nexth_loop}
        \If{$\neg l<h$} \Return{$l$}\EndIf
        \State{\Call{swap}{$a[l],a[h]$}}
        \State{${+}{+}l$}
      \EndWhile
    \EndFunction

    \Function{partition\_pivot}{$a,l,h$}
      \State{$p\gets \Call{median\_of\_three}{a,l+1,\lfloor h-l\rfloor/2,h-1}$} \label{line:partition:pivot_selection}
      \State{\Call{swap}{$a[l],a[p]$}}
      \State{\Return{\Call{partition}{a,l+1,h,l}}}
    \EndFunction
  \end{algorithmic}
  \caption{Partitioning}\label{alg:partition}
  \end{algorithm}

  Note that this function contains a subtle optimization: The loops in lines~\ref{partition:nextl_loop} and \ref{partition:nexth_loop}
  will iterate until they find a mismatched element. This iteration will not exceed the array bounds, only because the median selection
  placed \emph{stopper elements} at the beginning and end of the array. In particular when the element comparison function is
  cheap (e.g.\ for integer arrays), omitting an extra bounds check in the very inner loop of the sorting procedure can make a difference in practice.

  \subsection{Unguarded Insertion Sort}
  Insertion sort is a well-known simple sorting algorithm, which works by inserting the elements of the
  array, one after the other, into the sorted prefix of the array. Its worst case complexity is quadratic in general,
  but linear for partially sorted arrays with a constant threshold. For small thresholds, it performs better than quicksort,
  such that it is effective to stop the quicksort procedure early, and do the final sorting with insertion sort.
  Algorithm~\ref{alg:insort} shows the algorithm we implemented\footnote{The guarded
  version ($G=\textrm{true}$) slightly differs from the implementation in the GNU C++ Library. However, as it is only called for a small
  number of elements, we don't expect this to have any significant effect.}.

  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{insert}{$G,a,l,i$}
      \State{$t\gets a[i]$}
      \While{$ (\neg G \vee l<i) \wedge t<a[i-1]$} \label{alg:insert:loophead}
        \State{$a[i]\gets a[i-1]$}
        \State{${-}{-}i$}
      \EndWhile
      \State{$a[i]\gets t$}
    \EndProcedure

    \Procedure{gen\_insort}{$G,a,l,i,h$}
      \While{$i<h$}
        \State{\Call{insert}{$G,a,l,i$}}
        \State{${+}{+}i$}
      \EndWhile
    \EndProcedure

    \Procedure{final\_insort}{$a,l,h$}
      \If{$h-l\le\textrm{threshold}$}
        \State{\Call{gen\_insort}{$\textrm{true},a,l,l+1,h$}}
      \Else
        \State{\Call{gen\_insort}{$\textrm{true},a,l,l+1,l+\textrm{threshold}$}}
        \State{\Call{gen\_insort}{$\textrm{false},a,l,l+\textrm{threshold},h$}}
      \EndIf
    \EndProcedure
  \end{algorithmic}
  \caption{Final Insertion Sort}\label{alg:insort}
  \end{algorithm}

  This algorithm contains a similar optimization as the partitioning algorithm: As the array is already partially sorted,
  we know that elements beyond the threshold index won't be inserted at the very beginning of the array.
  This is exploited to get rid of the bounds check in line~\ref{alg:insert:loophead}.
  Only the first $\textrm{threshold}$ elements of the array are sorted with activated bounds check. In the displayed code, we control
  the bounds check by the Boolean parameter $G$, and assume that the algorithm will be specialized for both values of $G$.

%   Combination of quicksort, heapsort, and insertion sort
%   Worst-case $O(n\log n)$, average case efficiency of quicksort.

\section{Notation}
  Throughout this paper, we use some specific notation. We write function application in curried form, i.e., \is{f x_1 ... x$_n$}.
  Definitions of types and functions are denoted by \is{==}. Indexes start at zero. The \is$i$th element of a list is \is$xs[i]$,
  the slice of a list from index \is$i$ inclusive to \is$j$ exclusive is \is$xs[i..<j]$. We also use the notations \is{xs[i<..<j]}, \is{xs[i<..j]},
  and \is{xs[i..j]} to specify exclusiveness of bounds. Moreover, an omitted lower bound denotes index $0$, and an omitted upper bound denotes
  the length of the list, e.g., \is$xs[..<n]$ is the list of the first \is$n$ elements of $xs$, and \is$xs[n..]$ is the list of the remaining elements.
  List concatenation is written as \is{xs_1@xs_2}, and \is$xs[i:=x]$ is the list \is$xs$, with index \is$i$ updated to element \is$x$.
  The length of list \is$xs$ is denoted by \is{|xs|}.

\section{The Isabelle Refinement Framework}
  The Isabelle Refinement Framework~\cite{LaTu12} provides tools for
  verified program development by stepwise refinement, using the Isabelle HOL proof assistant~\cite{NPW02}.
  It is based on a nondeterminism-error monad, defined as follows:
  \begin{lstlisting}
    \<alpha> nres == FAIL | RES (\<alpha> set)
    return x == RES {x}
    bind FAIL f == FAIL | bind (RES X) f = |_|x\<in>X. f x
  \end{lstlisting}
  Intuitively, \is{FAIL} represents a program that may fail (e.g.~not terminate or fail explicitly by violating an assertion),
  and \is{RES X} represents a program that, nondeterministically, returns a value \is{x\<in>X}.
  By extending the subset ordering, with \is{FAIL} as greatest element, \is{\<alpha> nres} becomes a complete lattice.
  The program \is{return x} returns exactly the value \is{x}.
  The program \is{bind m f} sequentially composes \is{m} and \is{f}: it nondeterministically chooses a
  return value \is{x} of \is{m}, and then behaves like \is{f x}. If either \is{m}, or one of the \is{f x} may fail,
  then also \is{bind m f} may fail.

  We use sequential notation for bind, i.e., \is$x<-m; f x$ stands for \is$bind m (\<lambda>x. f x)$,
  and \is$m_1; m_2$ stands for \is$bind m_1 (\<lambda>_. m_2)$, i.e., the result of \is$m_1$ is ignored.
  We also use the standard HOL functions \is{if-then-else} and \is{let}, and define the following notations:
  \begin{lstlisting}
    assert \Phi == if \Phi then return () else FAIL
    spec x. \Phi x == RES {x | \Phi x}
  \end{lstlisting}
  An assertion will fail if its argument is false, and \is{spec x. \Phi x} describes all results that satisfy \is{\Phi}.

  Recursion is defined by the greatest fixed point over a monotonic functor \is{F}{\hspace{1pt}}\footnote{Monadic
    programs are monotonic by construction~\cite{Kr10}. Moreover, on monadic programs, the greatest fixed point over the \is{nres}-ordering
    coincides with the fixed point over a flat CCPO, which is the standard construction for recursion. In the actual
    formalization, monotonicity wrt.\ both orderings is asserted. }
  , and while loops are defined via recursion:
  \begin{lstlisting}
    rec F x == assert (mono F); fixp F x
    while b c s == rec (\<lambda>W s.
      if b s then s' <- c s; W s' else return s) s
  \end{lstlisting}
  From now on, we will omit the \is{rec} syntax, and define recursive monadic functions by their recursion equation, e.g.:
  \begin{lstlisting}
    while b c s ==
      if b s then s' <- c s; while b c s' else return s
  \end{lstlisting}

  The ordering on \is{nres} can be interpreted as refinement: \is{m_1 <= m_2} means
  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
  The statement
  \begin{lstlisting}
    f x <= { assert (P x); spec r. Q r}
  \end{lstlisting}
  means that function $f$ satisfies the specification with precondition \is{P} and postcondition \is{Q}.

\section{Verifying Introsort}
  The first step to verify a sorting algorithm is to specify the desired result.
  We specify a sorting algorithm as follows:
  \begin{lstlisting}
    sort_spec xs == spec xs'. sorted xs' \<and> mset xs' = mset xs
    $\textrm{where}$
    sorted xs == \<forall>i<j<|xs|. \<not>(xs[i] > xs[j])
  \end{lstlisting}
  That is, the result will be sorted and a permutation of the input (\is{mset xs} denotes the multiset of elements of list $xs$).

  Note that \is{<} is a \emph{strict weak ordering}, i.e., a relation with:
  \begin{lstlisting}
    a<b ==> \<not>b<a      and      a<c ==> a<b \or b<c
  \end{lstlisting}
  Weak orderings are used by C++ for sorting~\cite{Josu12}.
  They are a natural model for objects that are sorted by linearly ordered keys:
  An equivalent characterization can be given by an equivalence relation on the elements (e.g.~objects with same keys),
  and a linear ordering on the equivalence classes.
  In particular, any linear ordering is a weak ordering, with the equivalence relation being equality, and
  $\neg (a<b)$ meaning $b\le a$. While our formalization is over arbitrary weak orderings, we assume a linear ordering for the rest of
  this paper, to simplify the presentation.

  For a partial sorting algorithm wrt.\ threshold \is{n}, we replace \is{sorted} by:
  \begin{lstlisting}
    part_sorted xs == \<exists>ss. is_slicing xs ss \<and> sorted$_\preceq$ ss"
    $\textrm{where}$
    is_slicing xs ss == xs = concat ss \<and> (\<forall>s\<in>ss. |s| <= n)"
    xs $\preceq$ ys == \<forall>x\<in>xs. \<forall>y\<in>ys. x <= y
  \end{lstlisting}
  That is, a list is partially sorted if we can split it into slices that are not longer than
  the threshold $n$, and each element in a slice is less than or equal to any element in a subsequent slice.

>>>>

% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{balance}

\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}

@@

\usepackage[numbers]{natbib}

@@

\usepackage{pgfplots}

@@

  \newcommand{\false}{\textrm{false}}
  \newcommand{\true}{\textrm{true}}

% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]

@@

\title{Efficient Verified Implementation of Introsort and Pdqsort}

\author{Peter Lammich}
\institute{The University of Manchester, UK}
% \email{peter.lammich@machester.ac.uk}
%
% \authorinfo{Peter Lammich}
%            {The University of Manchester, UK}
%            {peter.lammich@machester.ac.uk}

@@

\begin{abstract}
  Sorting algorithms are an important part of most standard libraries, and
  both, their correctness and efficiency is crucial for many applications.

  As generic sorting algorithm, the GNU C++ Standard Library implements the Introsort algorithm,
  a combination of quicksort, heapsort, and insertion sort. The Boost C++ Libraries implement Pdqsort,
  an extension of Introsort that achieves linear runtime on inputs with certain patterns.

  We verify Introsort and Pdqsort in the Isabelle LLVM verification framework,
  closely following the state-of-the-art implementations from GNU and Boost.
  On an extensive benchmark set, our verified implementations perform on par with the originals.
\end{abstract}

@@

\section{Introduction}
%   In this paper, we present verified and efficient implementations of
%   Musser's introspective sorting algorithm (Introsort)~\cite{Muss97} and its
%   extension \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}.
%   Using the Isabelle Refinement Framework with its LLVM backend~\cite{La19-llvm},
%   we closely follow the implementations of Introsort and Pdqsort from
%   the GNU C++ Library (libstdc++)~\cite{libstdc++} and the Boost C++ Libraries~\cite{boost}.
%   The resulting sorting algorithms
%   are verified down to their representation in the LLVM intermediate language~\cite{LLVM-manual},
%   and perform on-par with the unverified original implementations on an extensive set of benchmarks.

  Sorting algorithms are an important part of any standard library.
  The GNU C++ Library (libstdc++)~\cite{libstdc++} implements Musser's introspective sorting algorithm (Introsort)~\cite{Muss97}.
  It is a combination of quicksort, heapsort, and insertion sort,
  which has the fast average case runtime of quicksort and the optimal $O(n\log(n))$
  worst-case runtime of heapsort.
  The Boost C++ Libraries~\cite{boost} provide a state-of-the-art implementation of
  \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}, an extension of Introsort to
  achieve better performance on inputs that contain certain patterns like already sorted sequences.
  Verification of these algorithms and their state-of-the-art implementations is far from trivial,
  but turns out to be manageable when handled with adequate tools.

  Sorting algorithms in standard libraries have not always been correct.
  The Timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
  Also, many real-world mergesort implementations suffered from an overflow bug~\cite{bs_flaw_blogpost}.
  Finally, LLVM's libc++~\cite{libc++} implements a different quicksort based sorting algorithm. While it may be functionally correct,
  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by January 2020.}.

  In this paper, we present efficient implementations of Introsort and Pdqsort that are verified down to their
  LLVM intermediate representation~\cite{LLVM-manual}. The verification uses the
  Isabelle Refinement Framework~\cite{LaTu12}, and its recent Isabelle-LLVM backend~\cite{La19-llvm}.
  We also report on two extensions of Isabelle-LLVM, to handle nested container data structures and to automatically
  generate C-header files to interface the generated code. Thanks to the modularity of the Isabelle Refinement Framework,
  our verified algorithms can easily be reused in larger verification projects.

  While sorting algorithms are a standard benchmark for theorem provers and program verification tools,
  verified real-world implementations seem to be rare: apart from our work, we are only aware of
  two verified sorting algorithms~\cite{GRBBH15,BSSU17} from the Java standard library.

  The complete Isabelle/HOL formalization and the benchmarks are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm/}.

@@

\section{The Introsort and Pdqsort Algorithms}
  The Introsort algorithm by Musser~\cite{Muss97} is a generic unstable sorting algorithm
  that combines the good average-case runtime of quicksort~\cite{Hoare61} with the optimal $O(n\log(n))$ worst-case complexity of heapsort~\cite{Will64}.
  The basic idea is to use quicksort as main sorting algorithm, insertion sort for small partitions, and heapsort when the recursion
  depth exceeds a given limit, usually $2\lfloor\log_2 n\rfloor$ for $n$ elements.

  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{introsort}{$xs,l,h$}
      \If{$h-l>1$} \label{l:intrs:trivial}
        \State{\Call{introsort\_aux}{$xs,l,h,2\lfloor\log_2(h-l)\rfloor$}} \label{l:intrs:aux}
        \State{\Call{final\_insort}{$xs,l,h$}} \label{l:intrs:finalis}
      \EndIf
    \EndProcedure

    \Procedure{introsort\_aux}{$xs,l,h,d$}
      \If{$h-l>\textrm{threshold}$} \label{l:intrs:threshold}
        \If{$d=0$}~\Call{heapsort}{$xs,l,h$} \label{l:intrs:heapsort}
        \Else
          \State{$m\gets\Call{partition\_pivot}{xs,l,h}$} \label{l:intrs:part}
          \State{\Call{introsort\_aux}{$xs,l,m,d-1$}}  \label{l:intrs:rec1}
          \State{\Call{introsort\_aux}{$xs,m,h,d-1$}}  \label{l:intrs:rec2}
        \EndIf
      \EndIf
    \EndProcedure
  \end{algorithmic}
  \caption{Introsort}\label{alg:introsort}
  \end{algorithm}

  Algorithm~\ref{alg:introsort} shows our implementation of Introsort, which closely follows the implementation in libstdc++~\cite{libstdc++}.
  The function \textsc{introsort} sorts the slice \is$l..<h$ of the list\footnote{%
    Our formalization initially uses lists to represent the sequence of elements to be sorted, and refines them to arrays later (cf.\ Sec.\ \ref{sec:imp_impl}).
  } $xs$.
  If there is more than one element (line~\ref{l:intrs:trivial}), it initializes a depth
  counter and calls the function \textsc{introsort\_aux} (line~\ref{l:intrs:aux}), which partially sorts the list such that every element is no more than
  threshold positions away from its final position in the sorted list. The remaining sorting is then done by insertion sort (line~\ref{l:intrs:finalis}).
  The function \textsc{introsort\_aux} implements a recursive quicksort scheme: recursion stops if the slice becomes smaller than the the threshold (line~\ref{l:intrs:threshold}).
  If the maximum recursion depth is exhausted, heapsort is used to sort the slice (line~\ref{l:intrs:heapsort}).
  Otherwise, the slice is partitioned (line~\ref{l:intrs:part}), and the procedure is
  recursively invoked for the two partitions (line~\ref{l:intrs:rec1}--\ref{l:intrs:rec2}).

  Note that we do not try to invent our own implementation, but closely follow the existing (and hopefully well-thought) libstdc++ implementation.
  This includes the slightly idiosyncratic partitioning scheme, which leaves the pivot-element as first element of the left partition.
  Moreover, the libstdc++ implementation contains a manual tail-call optimization, replacing the recursive call in line~\ref{l:intrs:rec2} by a loop.
  While we could easily add this optimization in an additional refinement step, it turned out to be unnecessary, as LLVM recognizes and
  eliminates this tail call automatically.

  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{pdqsort}{$lm,xs,l,h,d$}
      \If{$h-l>1$}~\Call{pdqsort\_aux}{$\true,xs,l,h,\log(h-l)$}\EndIf \label{l:pdq:aux}
    \EndProcedure

    \Procedure{pdqsort\_aux}{$lm,xs,l,h,d$}
      \If{$h-l<\textrm{threshold}$}~\Call{insort}{$lm,xs,l,h$} \label{l:pdq:threshold}
      \Else
        \State{\Call{pivot\_to\_front}{$xs,l,h$}} \label{l:pdq:pivotfront}
        \If{$\neg lm \wedge xs[l-1] \not< xs[l]$} \label{l:pdq:eqopt}
          \State{$m\gets \Call{partition\_left}{xs,l,h}$} \label{l:pdq:partleft}
          \State{\textbf{assert} $m+1\le h$} \label{l:pdq:assert}
          \State{\Call{pdqsort\_aux}{$\false,xs,m+1,h,d$}} \label{l:pdq:recright}
        \Else
          \State{$(m,ap)\gets\Call{partition\_right}{xs,l,h}$} \label{l:pdq:part}
          \If{$m-l < \lfloor(h-l)/8\rfloor \vee h-m-1 < \lfloor(h-l)/8\rfloor$} \label{l:pdq:unbal}
            \If{${-}{-}d = 0$}~\Call{heapsort}{xs,l,h};~\Return\EndIf \label{l:pdq:heapsort}
            \State{\Call{shuffle}{xs,l,h,m}} \label{l:pdq:shuffle}
          \ElsIf{$ap \wedge \Call{maybe\_sort}{xs,l,m} \wedge \Call{maybe\_sort}{xs,m+1,h}$}\State{\Return} \label{l:pdq:maybe}
          \EndIf
          \State{\Call{pdqsort\_aux}{$lm,xs,l,m,d$}}        \label{l:pdq:rec1}
          \State{\Call{pdqsort\_aux}{$\false,xs,m+1,h,d$}}  \label{l:pdq:rec2}
        \EndIf
      \EndIf
    \EndProcedure
  \end{algorithmic}
  \caption{Pdqsort}\label{alg:pdqsort}
  \end{algorithm}

  Algorithm~\ref{alg:pdqsort} shows our implementation of Pdqsort.
  As for Introsort, the wrapper \textsc{pdqsort} just initializes a depth counter, and then calls the function \textsc{pdqsort\_aux}
  (line~\ref{l:pdq:aux}),
  which, in contrast to Introsort, completely sorts the list, such that no final insertion sort is necessary.
  Again, the \textsc{pdqsort\_aux} function implements a recursive quicksort scheme, however, with a few additional optimizations.
  Slices smaller than the threshold are sorted with insertion sort (line~\ref{l:pdq:threshold}). If the current slice is not the leftmost one
  of the list, as indicated by the parameter $lm$,
  the element before the start of the slice is guaranteed to be smaller than any element of the slice itself. This can be exploited to omit a comparison
  in the inner loop of insertion sort (cf.~Sec.~\ref{sec:insort}).
  If the slice is not smaller than the threshold, a pivot element is selected and moved to the front of the slice (line~\ref{l:pdq:pivotfront}).
  If the pivot is equal to the element before the current slice (line~\ref{l:pdq:eqopt}), this indicates a lot of equal elements.
  The \textsc{partition\_left} function (line~\ref{l:pdq:partleft}) will put them in the left partition, and then only the right partition needs
  to be sorted recursively (line~\ref{l:pdq:recright}).
  Otherwise, \textsc{partition\_right} (line~\ref{l:pdq:part}) places elements equal to the pivot in the right partition.
  Additionally, it returns a flag $ap$ that indicates that the slice was already partitioned.
  Next, we check for a highly unbalanced partitioning (line~\ref{l:pdq:unbal}), i.e., if one partition is less than 1/8th of the overall size.
  After encountering a certain number of highly unbalanced partitionings, Pdqsort switches to heapsort (line~\ref{l:pdq:heapsort}).
  Otherwise, it will shuffle some elements in both partitions, trying to break up patterns in the input (line~\ref{l:pdq:shuffle}).
  If the input was already partitioned wrt.\ the selected pivot (indicated by the flag $ap$),
  Pdqsort will optimistically try to sort both partitions with insertion sort (line~\ref{l:pdq:maybe}).
  However, these insertion sorts abort if they cannot sort the list with a small number of swaps, limiting the penalty for being too optimistic.
  Finally, the two partitions are recursively sorted (lines~\ref{l:pdq:rec1}--\ref{l:pdq:rec2}).

  Our implementation of Pdqsort closely follows the implementation we found in Boost~\cite{boost}.
  Again, we omitted a manual tail call optimization that LLVM does automatically.
  Moreover, for certain comparison functions, Boost's Pdqsort uses a special branch-aware partitioning algorithm~\cite{EdWe16}.
  We leave its verification to future work, but note that it will easily integrate in our existing formalization.

  While Introsort and Pdqsort are based on the same idea, this presentation focuses on the more complex Pdqsort:
  apart from the more involved \textsc{pdqsort\_aux} function,
  \textsc{pivot\_to\_front} uses Tukey's 'ninther' pivot selection~\cite{Bent93}, while Introsort uses the simpler median-of-three scheme.
  It has two partitioning algorithms used in different situations, and the \textsc{partition\_right} algorithm also
  checks for already partitioned slices. Finally, with \textsc{insort} and \textsc{maybe\_sort}, it uses two different
  versions of insertion sort.

  \section{Verification}
  We use the Isabelle Refinement Framework~\cite{LaTu12,La19-llvm} to formally verify our algorithms.
  It provides tools to develop algorithms by stepwise refinement, and generates code in the LLVM intermediate representation~\cite{LLVM-manual}.

  A program returns an element of the following datatype:
  \begin{lstlisting}
    \<alpha> nres == fail | spec (\<alpha> => bool)
  \end{lstlisting}
  Here \is{fail} represents possible non-termination or assertion violation, and \is{spec P}
  a result nondeterministically chosen to satisfy predicate $P$. Note that we use \is{==} to indicate defining equations.
  We define a \emph{refinement ordering} on \is{nres} by
  \begin{lstlisting}
    spec P <= spec Q == \<forall>x. P x ==> Q x         fail $\not\le$ spec Q          m <= fail
  \end{lstlisting}
  Intuitively, \is{m_1 <= m_2} means
  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
  Note that \is$<=$ is a complete lattice, with top element \is$fail$.
  The \emph{monad combinators} are then defined as
  \begin{lstlisting}
    return x == spec y. y=x
    bind (spec P) f == |_|{f x | P x}         bind fail f == fail
  \end{lstlisting}
  Here, \is{return x} deterministically returns \is{x}, and \is{bind m f} chooses a result of \is$m$ and then applies
  \is$f$ to it. If \is$m$ may fail, then the bind may also fail.
  We write \is{ x<-m; f x} for \is{bind m (\<lambda>x. f x)}, and \is{m_1; m_2} for \is{bind m_1 (\<lambda>_. m_2)}.

%   Programs are expressed in the following nondeterminism-error monad.
%   \begin{lstlisting}
%     \<alpha> nres == fail | spec (\<alpha> => bool)                return x == spec y. y=x
%     bind fail f == fail | bind (spec P) f = |_|{f x | P x}
%   \end{lstlisting}
%   Here \is{fail} represents possible non-termination or assertion violation, and \is{spec P}
%   a result nondeterministically chosen to satisfy predicate $P$.
%   We define a \emph{refinement ordering} on \is{nres} by
%   \begin{lstlisting}
%     spec P <= spec Q == \<forall>x. P x --> Q x         fail $\not\le$ spec Q          m <= fail
%   \end{lstlisting}
%   Intuitively, \is{m_1 <= m_2} means
%   that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
%   Note that \is$<=$ is a complete lattice, with top element \is$fail$.
%   The program \is{return x} returns exactly the value \is{x}.
%   The program \is{bind m f} chooses a result of \is$m$ and then applies \is$f$ to it. If \is$m$ may fail, then the bind may also fail.
%   We will write \is{ x<-m; f x} for \is{bind m (\<lambda>x. f x)}, and \is{m_1; m_2} for \is{bind m_1 (\<lambda>_. m_2)}.

  Arbitrary recursive programs can be defined via a fixed-point construction~\cite{Kr10}.
  An \emph{assertion} fails if its condition is not met, otherwise it returns the unit value:
  \begin{lstlisting}
    assert P == if P then return () else fail;
  \end{lstlisting}
  Assertions are used to express that a program \is{m} satisfies the Hoare triple with
  precondition \is$P$ and postcondition \is$Q$:
  \begin{lstlisting}
    m <= assert P; spec x. Q x
  \end{lstlisting}
  If the precondition is false, the right hand side is \is{fail}, and the statement trivially holds.
  Otherwise, $m$ cannot fail, and every possible result $x$ of $m$ must satisfy $Q$.

  While the Isabelle Refinement Framework provides some syntax to express programs, for better
  readability, we use the slightly more informal syntax that we have already used in Algorithms~\ref{alg:introsort} and \ref{alg:pdqsort}.
  In particular, we treat lists as if they were updated in place, while our actual formalization is purely functional, i.e., generates a new
  version of the list on each update, which is explicitly threaded through the program. Destructively updated arrays will only be introduced in a
  later refinement step (cf.~Sec.~\ref{sec:imp_impl}).

  \subsection{Specification of Sorting Algorithms}\label{sec:sortspec}
  The first step to verify a sorting algorithm is to specify the desired result.
  We specify a sorting algorithm as follows:
  \begin{lstlisting}
    sort_spec xs l h
      == assert l<=h \<and> h<=|xs|; spec xs'. xs =$_{l,h}$ xs' \<and> sorted (xs'[l..<h])
  \end{lstlisting}
%     xs =$_{l,h}$ xs'
%       == xs[0..<l]=xs'[0..<l] \<and> mset (xs[l..<h]) = mset (xs'[l..<h]) \<and> xs[h..]=xs'[h..]
  here \is$|xs|$ is the length of the list \is{xs} and
  \is{xs[I]} is the slice of the list \is{xs} for indexes in the interval \is{I}.
%   and \is{xs[l..]} is short for \is{xs[l..<|xs|${}$]}.
  The equivalence relation \is{xs =$_{l,h}$ xs'} relates lists $xs$ and $xs'$ iff they are equal outside the slice \is$l..<h$
  and $xs$ is a permutation of $xs'$.
  To simplify the presentation, we assume a linear ordering on the elements.
  Note that both C++ and our actual formalization support arbitrary weak orderings~\cite{Josu12}.
